FGFR1/3 Signaling as Achilles’ Heel of Phenotypic Diversity in Urothelial Carcinoma

### 2025-04-28 report/markdown/supplementary_material.Rmd
Figure/Table/Supplement reshaping
Changes in composition of figures and tables requested by a reviewer. The whole content of tables is now in Supplementary Material. Parts of Figure 1 were moved to Supplementary Material

For MIBC molecular class assignment and machine learning modeling tasks, $log_2$-transformed gene expression levels were processed by ComBat (R package _sva_ and [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET)) [@Leek2012] to correct for the inter-cohort variability. 
Assignment of the TCGA BLCA, IMvigor, and BCAN cancer samples to consensus molecular classes was done with the nearest centroid algorithm fed with ComBat-normalized whole-genome $log_2$ mRNA or Z-score expression levels (R package _consensusMIBC_) [@Kamoun2020]. 
Specimens without unequivocal assignment ('not-assigned') were excluded from further analyses. 
Please note that due to insufficient numbers of specimens assigned to the 
luminal non-specified (LumNS; `r subtypes$assignment %>% map(filter, consensusClass == 'LumNS') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')`samples) and 
neuroendocrine-like classes (NE-like; `r subtypes$assignment %>% map(filter, consensusClass == 'NE-like') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')` samples), 
we restricted our analyses to the major classes, i.e. 
luminal papillary (LumP; `r subtypes$assignment %>% map(filter, consensusClass == 'LumP') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')` samples), 
luminal unstable (LumU; `r subtypes$assignment %>% map(filter, consensusClass == 'LumU') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')`), 
stroma-rich (`r subtypes$assignment %>% map(filter, consensusClass == 'Stroma-rich') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')`), and 
basal/squamous-like classes (Ba/Sq; `r subtypes$assignment %>% map(filter, consensusClass == 'Ba/Sq') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')`). 
Demographic, clinical, and pathological characteristic of the consensus molecular classes of MIBC is presented in __Supplementary Table S\@ref(tab:tab-sub-clinic)__.

@@

Predictions of resistance to pan-FGFR inhibitors in the TCGA BLCA, IMvigor, and BCAN transcriptomic cohorts in form of IC50 (drug concentration resulting in 50% growth inhibition, GDSC1, GDSC2, CTRP2) and AUC (area under the dose-response curve, PRISM) were made by RIDGE regularized linear models trained with the drug response data and baseline transcriptome of epithelial cancer cell lines in GDSC1, GDSC2, CTRP2, and PRISM screening experiments [@Yang2013; @Seashore-Ludlow2015; @Corsello2020]. 
The general idea of modeling drug resistance in bulk cancer transcriptome samples by RIDGE models trained with in vitro drug response data was proposed by Geeleher and colleagues [@Geeleher2014] and used in our recent publication on testicular cancer [@Torzsok2025]. 
The models were trained with ComBat-processed $log_2$ baseline expression values of gene shared by the cell line and bulk cancer data sets. 
The IC50 values were $log$-transformed to improve normality. 
The pre-processing of expression and in vitro drug response data was accomplished by `multi_process()` function from [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET) package. 
Model tuning, i.e. choice of the shrinkage parameter $\lambda$ based on the minimal mean squared error in 10-fold cross-validation, and training in the cell line data sets was accomplished by function `train()` from [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET) package. 
The RIDGE models of IC50 and AUC characterized by cross-validated $R^2 \geq$ 0.13 and correlation between the predicted and observed drug response with Pearson's r $\geq$ 0.5 were subsequently used for predictions of the resistance for bulk cancer transcriptome samples. 
Those criteria were fulfilled for `r drugs$fgfr_drugs %>% length` pan-FGFR inhibitors, i.e. inhibitors targeting at least FGFR1/2/3. 
Those substances were: `r list(x = drugs$fgfr_drugs %>% exchange(drugs$lexicon), y = drugs$fgfr_drugs %>% exchange(drugs$lexicon, value = 'experiment'), z = drugs$fgfr_drugs %>% exchange(drugs$lexicon, value = 'drug_id')) %>% pmap_chr(function(x, y, z) paste0(x, ' (', y, ', ID: ', z, ')')) %>% collapse_and`. 
The predictions of IC50 and AUC were made by calling `predict()` method for the object returned by `train()` function (package [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET)).

@@

Mutation status, baseline expression measured by RNA sequencing, and effect scores of CRISP/Cas9 and RNA interference of the FGFR-, FGF-, and FGFBP-coding genes in urothelial cancer cell lines were downloaded from the [DepMap](https://depmap.org/portal/) repository [@DepMap2024]. 
No information treatment of the cell line donors or pre-treatment of the cell lines was available. 
The available cell lines were `r map2_chr(depmap$cell_lines$cell_line_name, depmap$cell_lines$sample_id, ~paste0(.x, ' (ID:', .y, ')')) %>% collapse_and`. 
The mutation status was 0/1 coded, where 1 stood for presence of at least one somatic mutation in a gene. 
RNA sequencing information was provided as $log_2$ TPM. 
Resistance of urothelial cancer cell lines to pan-FGFR inhibitors and doxorubicin considered as a positive control was assessed for urothelial cancer cell lines in the GDSC1/2 (compounds: doxorubicin, `r depmap$gdsc_lexicon %>% filter(!drug_name %in% c('Doxorubicin', 'FGFR_0939'), !duplicated(drug_name)) %>% .$drug_name %>% collapse_and`, source [GDSC data portal](https://www.cancerrxgene.org/)) and PRISM experiments (compounds: doxorubicin, `r depmap$prism_lexicon %>% filter(drug_name != 'doxorubicin', !duplicated(drug_name)) %>% .$drug_name %>% collapse_and`, source: [DepMap](https://depmap.org/portal/)) [@Yu2016; @Yang2013]. 
The resistance was measured in the GDSC1/2 experiments by IC50; sensitive cell lines were defined by IC50 < 10 µM, which was the highest concentration of the FGFR inhibitors in the assays. 
The resistance in the PRISM assays was gauged by AUC normalized for effects of DMSO control; sensitive cell lines were considered for AUC < 1.

@@

Differences in $log_2$-transformed transcript levels of the genes of interest of ssGSEA scores of the FGFR-related gene signatures between the LumP, LumU, stroma-rich, and Ba/Sq MIBC consensus molecular classes of urothelial cancers were explored with a two step procedure. 
In the first step, differences of expression levels or ssGSEA scores between the analysis groups were assessed by one-way ANOVA with $\eta^2$ effect size statistic. 
In the second step, differences between a given class and the LumP class were were evaluated by Welch's two-tailed T test with Cohen's d effect size. 
The significantly differentially regulated genes were defined by pFDR(ANOVA) < 0.05, $\eta^2 \geq$ 0.14, pFDR(T test) < 0.05, and at least 1.25-fold up- or downregulation as compared with the baseline. 
The significantly differentially regulated gene signatures were defined by pFDR(ANOVA) < 0.05, $\eta^2 \geq$ 0.14, pFDR(T test) < 0.05, and Cohen's d $\geq$ 0.5. 
A similar testing procedure was applied to predictions of IC50 and AUC of pan-FGFR inhibitors. 
However, due to non-normal distribution of some of the resistance metrics, Kruskal-Wallis with with $\eta^2$ statistic for testing for overall differences between the consensus subsets, and Mann-Whitney post-hoc test with biserial r effect size statistic was used for comparisons versus LumP cancers. 
The significant differences in predicted resistance were considered for pFDR(Kruskal-Wallis) < 0.05, $\eta^2 \geq$ 0.14, pFDR(Mann-Whitney) < 0.05, and biserial r $\geq$ 0.3.

@@

The expression values were fed into the models in form of Z-scores of Combat-processed $log_2$ gene counts (TCGA BLCA, IMvigor) or Z-scores of $log_2$ gene counts (BCAN). 
Assignment to the LumP, LumU, stroma-rich, and Ba/Sq consensus molecular classes by the nearest centroid classifier of _consensusMIBC_ R package served as the modeling response and ground truth for evaluation of the models.

The models were developed in a training subset derived from randomly chosen two-thirds of TCGA BLCA observations (n = `r nrow(ml_globals$data$tcga_train)`). 
In order to improve recognition of the minority consensus classes, the LumU and Stroma-rich subsets, we amplified their observations by over-sampling with the SMOTE algorithm (function `smote()`, package _performanceEstimation_) [@Torgo2014; @Chawla2002; @Torgo2016]. 
The Elastic Net model of consensus molecular classes was optimized (tuned) and trained in the SMOTE-amplified training data set with a wrapper around functions `cv.glmnet()` and `glmnet()` from _glmnet_ package [@Friedman2010]. 
The optimal values of the regularization parameter $\lambda$ was found by minimizing the Elastic Net models' deviance in 200 repeats 10-fold cross-validation. 
The Random Forest models of consensus molecular subsets were tuned and trained inn the SMOTE-amplified training data set with a wrapper around `ranger()` function from _ranger_ package [@Wright2017]. 
The optimal set of hyper-parameters that control behavior of the Random Forest models, 
mtry (number of randomly chosen variables per tree learner), 
minimal node size (minimal number of observations in the terminal nodes of the tree learner), and 
splitting rule (rule that defines partitioning of the explanatory variables at the learner's nodes) 
was found by minimizing the out-of-bag classification error [@Breiman2001]. 
The final Random Forest models consisted of 500 tree learners. 
The optimal values of the tunable hyper-parameters are listed in __Supplementary Table S\@ref(tab:tab-tuning)__.

Performance of the machine learning models was evaluated by comparing the predicted and actual consensus molecular class in the unmodified test subset of cancers in the TCGA BLCA cohort (`r ml_globals$data$tcga_test %>% nrow`), as well as in the IMvigor (n = `r ml_globals$data$imvigor %>% nrow`) and BCAN (n = `r ml_globals$data$bcan %>% nrow`) independent validation cohorts. 
The consensus classification returned by the _consensusMIBC_ R package nearest centroid tool was assumed the ground truth. 
The following numeric statistics of model performance were computed: 
overall accuracy (fraction of cancer samples with correct consensus molecular class prediction), 
Cohen's $\kappa$ (non-accidental concordance between the predicted and actual class, high $\kappa$ expected for a highly accurate model) [@McHugh2012], 
Brier score (metric of overall model calibration, low Brier scores expected for a well calibrated model) [@Brier1950; @Goldstein-Greenwood2021], 
and receiver-operating characteristic measures such as area under the curve (AUC, measure of overall sensitivity and specificity). 
Those evaluation metrics were computed with `multiClassSummary()` function from _caret_ package [@Kuhn2008] and an in-house developed function that calculates Brier scores for the models of interest and a nonsense model predicting the classes at random (__Supplementary Figure S\@ref(fig:fig-ml-eval)A__, __Supplementary Table S\@ref(tab:tab-ml)__) 
Additionally, accuracy of prediction for single consensus molecular classes was assessed by inspection of the confusion matrices (__Supplementary Figure S\@ref(fig:fig-confusion)__). 
Calibration of the machine learning models at predictions of particular consensus molecular classes was assessed with squares of differences between the predicted class assignment probabilities and the binarized actual class assignment (1: the class of interest, 0: any other consensus molecular class) or 'Brier squares' between the consensus molecular classes (__Supplementary Figure S\@ref(fig:fig-ml-eval)B__).

Importance of the explanatory variables, i.e. FGFR-, FGF-, and FGFBP-coding transcripts, and their positive and negative association with predictions of LumP, LumU, stroma-rich, and Ba/Sq classes by the machine learning models were assessed with the SHAP algorithm (Shapley additive explanations) [@Lundberg2017; @Covert2020]. 
To this end, we re-fitted the Elastic Net and Random Forest models of consensus molecular classes in the unmodified test subset of the TCGA BLCA cohort not used for the model training with the optimal set of the tuning parameters. 
Matrices of SHAP metrics for the genes and observations were computed for the models with `kernelshap()` function from _kernelshap_ package [@Mayer2023; @Mayer2023a]. 
As the background for the SHAP procedure, we used randomly selected 30% of observations of the test subset of the TCGA BLCA collective. 
Mean absolute SHAP value served as a measure of gene importance (__Supplementary Table S\@ref(tab:tab-shap)__). 
SHAP metrics for single observations and the mean absolute SHAP values were visualized as violin/swarm and bar plots with tools of _shapviz_ package [@Mayer2023] and in-house developed R functions (__Figure 1F__, __Supplementary Figure S\@ref(fig:fig-shap-elnet)__ - __S\@ref(fig:fig-shap-fgfr)__).

@@

flextable::flextable(suppl_tabs$cohorts) %>% 
  set_widths(c(2.5, rep(2.3, 6), 2.3, 2)) %>% 
  footnote(1, 8:9, 
           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

flextable::flextable(suppl_tabs$sub_clinic) %>% 
  merge_v(1) %>% 
  set_widths(c(2.2, 2.5, rep(2.8, 4), 2.3, 2)) %>% 
  footnote(1, 7:8, 
           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

```{r tab-fgf-genetics, tab.cap = "Frequency of somatic mutations and copy number variants of FGF-, FGFR-, and FGFBP-coding genes in the investigated cohorts. Statistical significance for differences between the cohorts was assessed by chi-squared test with Cramer's V effect size statistic. P values were corrected for multiple testing with the false discovery rate method. The table is available as a supplementary Excel file."}

suppl_tabs$genetic_alterations[1, 1] %>% 
  mutate(`Alteration type` = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(18)

@@

```{r tab-sub-genetics, tab.cap = "Frequency of somatic mutations and copy number variants of FGFR-, FGF-, and FGFBP-coding genes in the consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts. The frequencies are presented as percentages of the consensus class. Statistical significance of differences between the consensus classes was determined by weighted permutation test with Cramer's V effect size statistic. P values were corrected for multiple testing with the false discovery rate method. The table is available as a supplementary Excel file."}

suppl_tabs$sub_genetics[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(18)

@@

\newepage

```{r tab-expression, tab.cap = "Expression of FGFR-, FGF-, and FGFBP-coding genes at mRNA level was quantified in the TCGA BLCA, IMvigor, and BCAN cohort by RNA sequencing. The mRNA amounts were provided as log2-transformed gene counts (TCGA BLCA and IMvigor) or Z-scores of log2 gene counts (BCAN). Protein expression of the genes of interest was measured in the Human Protein Atlas (HPA) by immunohistochemistry (IHC). The protein amounts are metered by the IHC score defined as a product of staining intensity and quantity of positive cells. Median mRNA and protein amounts with interquartile ranges, ranges, and numbers of cancer samples are listed. The table is available as a supplementary Excel file."}

suppl_tabs$expression[1, 1] %>% 
  mutate(Variable = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(18)

@@

```{r tab-sub-dge, tab.cap = "Expression of FGF-, FGFR-, and FGFBP-coding genes in the consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts. Statistical significance of differences between the molecular classes was determined by one-way ANOVA with eta-square effect size statistic. P values were corrected for multiple testing with the false discovery rate method. log2-transformed expression levels are presented as medians with interquartile ranges and ranges. The table is available as a supplementary Excel file."}

suppl_tabs$sub_dge[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(18)

@@

```{r tab-sub-sig, tab.cap = "Single sample gene set enrichment analysis (ssGSEA) scores of FGFR-related gene signatures in the consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts. Statistical significance of differences between the molecular classes was determined by one-way ANOVA with eta-square effect size statistic. P values were corrected for multiple testing with the false discovery rate method. ssGSEA scores are presented as medians with interquartile ranges and ranges. The table is available as a supplementary Excel file."}

suppl_tabs$sub_sig[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(18)

@@

```{r tab-shap, tab.cap = "Importance of explanatory variables for the Elastic Net and Random Forest models of MIBC consensus molecular classes estimated by the SHAP algorithm. Contribution of the genuine MIBC consensus class-defining genes and the FGFR-, FGF-, and FGFBP-coding genes to prediction of consensus molecular classes was assessed by mean absolute values of SHAP metrics (Shapley additive explanations). The table is available as a supplementary Excel file."}

>>>>

For MIBC molecular class assignment and machine learning modeling tasks, $log_2$-transformed gene expression levels were processed by ComBat (R package _sva_ and [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET)) [@Leek2012] to correct for the inter-cohort variability. 
Assignment of the TCGA BLCA, IMvigor, and BCAN cancer samples to consensus molecular classes was done with the nearest centroid algorithm fed with ComBat-normalized whole-genome $log_2$ mRNA or Z-score expression levels (R package _consensusMIBC_) [@Kamoun2020]. 
Specimens without unequivocal assignment ('not-assigned') were excluded from further analyses. 
Please note that due to insufficient numbers of specimens assigned to the 
luminal non-specified (LumNS; `r subtypes$assignment %>% map(filter, consensusClass == 'LumNS') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')` samples) and 
neuroendocrine-like classes (NE-like; `r subtypes$assignment %>% map(filter, consensusClass == 'NE-like') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')` samples), 
we restricted our analyses to the major classes, i.e. 
luminal papillary (LumP; `r subtypes$assignment %>% map(filter, consensusClass == 'LumP') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')` samples), 
luminal unstable (LumU; `r subtypes$assignment %>% map(filter, consensusClass == 'LumU') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')` samples), 
stroma-rich (`r subtypes$assignment %>% map(filter, consensusClass == 'Stroma-rich') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')` samples), and 
basal/squamous-like classes (Ba/Sq; `r subtypes$assignment %>% map(filter, consensusClass == 'Ba/Sq') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')` samples). 
Demographic, clinical, and pathological characteristic of the consensus molecular classes of MIBC is presented in __Supplementary Table S\@ref(tab:tab-sub-clinic)__.

@@

Predictions of resistance to pan-FGFR inhibitors in the TCGA BLCA, IMvigor, and BCAN transcriptomic cohorts in form of IC50 (drug concentration resulting in 50% growth inhibition, GDSC1, GDSC2, CTRP2) and AUC (area under the dose-response curve, PRISM) were made by RIDGE regularized linear models trained with the drug response data and baseline transcriptome of epithelial cancer cell lines in GDSC1, GDSC2, CTRP2, and PRISM screening experiments [@Yang2013; @Seashore-Ludlow2015; @Corsello2020]. 
The general idea of modeling drug resistance in bulk cancer transcriptome samples by RIDGE models trained with in vitro drug response data was proposed by Geeleher and colleagues [@Geeleher2014] and used in our recent publication on testicular cancer [@Torzsok2025]. 
The models were trained with ComBat-processed $log_2$ baseline expression values of gene shared by the cell line and bulk cancer data sets. 
The IC50 values were $log$-transformed to improve normality. 
The pre-processing of expression and in vitro drug response data was accomplished by `multi_process()` function from [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET) package. 
Model tuning, i.e. choice of the shrinkage parameter $\lambda$ based on the minimal mean squared error in 10-fold cross-validation, and training in the cell line data sets was accomplished by function `train()` from [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET) package. 
The RIDGE models of IC50 and AUC characterized by cross-validated $R^2 \geq$ 0.13 and correlation between the predicted and observed drug response with Pearson's r $\geq$ 0.5 were subsequently used for predictions of the resistance for bulk cancer transcriptome samples. 
Those criteria were fulfilled for `r drugs$fgfr_drugs %>% length` pan-FGFR inhibitors, i.e. inhibitors targeting at least FGFR1/2/3. 
Those substances were: `r list(x = drugs$fgfr_drugs %>% exchange(drugs$lexicon), y = drugs$fgfr_drugs %>% exchange(drugs$lexicon, value = 'experiment'), z = drugs$fgfr_drugs %>% exchange(drugs$lexicon, value = 'drug_id')) %>% pmap_chr(function(x, y, z) paste0(x, ' (', y, ', ID: ', z, ')')) %>% collapse_and`. 
The predictions of IC50 and AUC were made by calling `predict()` method for the model container object generated by `train()` function (package [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET)).

@@

Mutation status, baseline expression measured by RNA sequencing, and effect scores of CRISP/Cas9 and RNA interference of the FGFR-, FGF-, and FGFBP-coding genes in urothelial cancer cell lines were downloaded from the [DepMap](https://depmap.org/portal/) repository [@DepMap2024]. 
No information treatment of the cell line donors or pre-treatment of the cell lines was available. 
The available cell lines were `r map2_chr(depmap$cell_lines$cell_line_name, depmap$cell_lines$sample_id, ~paste0(.x, ' (ID:', .y, ')')) %>% collapse_and`. 
The mutation status was 0/1 coded, where 1 stood for presence of at least one somatic mutation in a gene. 
RNA sequencing information was provided as $log_2$ TPM. 
Resistance of urothelial cancer cell lines to pan-FGFR inhibitors and doxorubicin considered as a positive cytotoxicity control was assessed for urothelial cancer cell lines in the GDSC1/2 (compounds: doxorubicin, `r depmap$gdsc_lexicon %>% filter(!drug_name %in% c('Doxorubicin', 'FGFR_0939'), !duplicated(drug_name)) %>% .$drug_name %>% collapse_and`, source [GDSC data portal](https://www.cancerrxgene.org/)) and PRISM experiments (compounds: doxorubicin, `r depmap$prism_lexicon %>% filter(drug_name != 'doxorubicin', !duplicated(drug_name)) %>% .$drug_name %>% collapse_and`, source: [DepMap](https://depmap.org/portal/)) [@Yu2016; @Yang2013]. 
The resistance was measured in the GDSC1/2 experiments by IC50; sensitive cell lines were defined by IC50 < 10 µM, which was the highest concentration of the FGFR inhibitors in the assays. 
The resistance in the PRISM assays was gauged by AUC normalized to the DMSO control; sensitive cell lines were considered for AUC < 1.

@@

Differences in $log_2$-transformed transcript levels of the genes of interest of ssGSEA scores of the FGFR-related gene signatures between the LumP, LumU, stroma-rich, and Ba/Sq consensus molecular classes of MIBC were explored with a two step procedure. 
In the first step, differences of the expression levels or ssGSEA scores between the analysis groups were assessed by one-way ANOVA with $\eta^2$ effect size statistic. 
In the second step, differences between a given class and the LumP class were were evaluated by Welch's two-tailed T test with Cohen's d effect size. 
The significantly differentially regulated genes were defined by pFDR(ANOVA) < 0.05, $\eta^2 \geq$ 0.14, pFDR(T test) < 0.05, and at least 1.25-fold up- or downregulation as compared with the baseline. 
The significantly differentially regulated gene signatures were defined by pFDR(ANOVA) < 0.05, $\eta^2 \geq$ 0.14, pFDR(T test) < 0.05, and Cohen's d $\geq$ 0.5. 
A similar testing procedure was applied to the predictions of IC50 and AUC of pan-FGFR inhibitors made for bulk cancer samples. 
However, due to non-normal distribution of some of the resistance metrics, Kruskal-Wallis with with $\eta^2$ statistic for testing for overall differences between the consensus subsets, and Mann-Whitney post-hoc test with biserial r effect size statistic was used for comparisons versus LumP cancers. 
Significant differences in the predicted resistance were considered for pFDR(Kruskal-Wallis) < 0.05, $\eta^2 \geq$ 0.14, pFDR(Mann-Whitney) < 0.05, and biserial r $\geq$ 0.3.

@@

The expression values were fed into the models in form of Z-scores of Combat-processed $log_2$ gene counts (TCGA BLCA, IMvigor) or Z-scores of $log_2$ gene counts (BCAN). 
Assignment to the LumP, LumU, stroma-rich, and Ba/Sq consensus molecular classes by the nearest centroid classifier of _consensusMIBC_ R package served as the modeling response and the ground truth for evaluation of the model predictions.

The models were developed in a training subset derived from randomly chosen two-thirds of TCGA BLCA observations (n = `r nrow(ml_globals$data$tcga_train)`). 
In order to improve recognition of the minority consensus classes, the LumU and Stroma-rich subsets, we amplified their observations by over-sampling with the SMOTE algorithm (function `smote()`, package _performanceEstimation_) [@Torgo2014; @Chawla2002; @Torgo2016]. 
The Elastic Net model of consensus molecular classes was optimized (tuned) and trained in the SMOTE-amplified training data set with a wrapper around functions `cv.glmnet()` and `glmnet()` from _glmnet_ package [@Friedman2010]. 
The optimal values of the regularization parameter $\lambda$ was found by minimizing the Elastic Net models' deviance in 200 repeats 10-fold cross-validation of the SMOTE augmented training data set. 
The Random Forest models of consensus molecular subsets were tuned and trained in the SMOTE-amplified training data set with a wrapper around `ranger()` function from _ranger_ package [@Wright2017]. 
The optimal set of hyper-parameters that control behavior of the Random Forest models, 
mtry (number of randomly chosen variables per tree learner), 
minimal node size (minimal number of observations in the terminal nodes of the tree learner), and 
splitting rule (rule that defines partitioning of the explanatory variables at the learner's nodes) 
was found by minimizing the out-of-bag classification error [@Breiman2001]. 
The final Random Forest models consisted of 500 tree learners. 
The optimal values of the tunable hyper-parameters are listed in __Supplementary Table S\@ref(tab:tab-tuning)__.

Performance of the machine learning models was evaluated by comparing the predicted and actual consensus molecular class in the unmodified test subset of cancers in the TCGA BLCA cohort (`r ml_globals$data$tcga_test %>% nrow`), as well as in the IMvigor (n = `r ml_globals$data$imvigor %>% nrow`) and BCAN (n = `r ml_globals$data$bcan %>% nrow`) independent validation cohorts. 
The consensus classification returned by the _consensusMIBC_ R package nearest centroid tool was assumed the ground truth. 
The following numeric statistics of model performance were computed: 
overall accuracy (fraction of cancer samples with correct consensus molecular class prediction), 
Cohen's $\kappa$ (non-accidental concordance between the predicted and actual class, high $\kappa$ expected for a highly accurate model) [@McHugh2012], 
Brier score (metric of overall model calibration, low Brier scores expected for a well calibrated model) [@Brier1950; @Goldstein-Greenwood2021], 
and receiver-operating characteristic measures such as area under the curve (AUC, measure of overall sensitivity and specificity). 
These evaluation metrics were computed with `multiClassSummary()` function from _caret_ package [@Kuhn2008] and an in-house developed function that calculates Brier scores for the models of interest and a nonsense model predicting the classes at random (__Supplementary Figure S\@ref(fig:fig-ml-eval)A__, __Supplementary Table S\@ref(tab:tab-ml)__) 
Additionally, accuracy of prediction for single consensus molecular classes was assessed by inspection of the confusion matrices (__Supplementary Figure S\@ref(fig:fig-confusion)__). 
Calibration of the machine learning models at predictions of particular consensus molecular classes was assessed with squares of differences between the predicted class assignment probabilities and the binarized actual class assignment (1: the class of interest, 0: any other consensus molecular class) or 'Brier squares' between the consensus molecular classes (__Supplementary Figure S\@ref(fig:fig-ml-eval)B__).

Importance of the explanatory variables, i.e. FGFR-, FGF-, and FGFBP-coding transcripts, and their positive and negative association with predictions of LumP, LumU, stroma-rich, and Ba/Sq classes by the machine learning models were assessed with the SHAP algorithm (Shapley additive explanations) [@Lundberg2017; @Covert2020]. 
To this end, we re-fitted the Elastic Net and Random Forest models of consensus molecular classes in the unmodified test subset of the TCGA BLCA cohort not used for the model training with the optimal set of the tuning parameters. 
Matrices of SHAP metrics for the genes and observations were computed for the models with `kernelshap()` function from _kernelshap_ package [@Mayer2023; @Mayer2023a]. 
As the background for the SHAP procedure, we used randomly selected 30% of observations of the test subset of the TCGA BLCA collective. 
Mean absolute SHAP value served as a measure of gene importance (__Supplementary Tables S\@ref(tab:tab-shap)__ and __S\@ref(tab:tab-kamoun-shap)__). 
SHAP metrics for single observations and the mean absolute SHAP values were visualized as violin/swarm and bar plots with tools of _shapviz_ package [@Mayer2023] and in-house developed R functions (__Figure 1D__, __Supplementary Figure S\@ref(fig:fig-rf-shap)__ - __S\@ref(fig:fig-shap-fgfr)__).

@@

flextable::flextable(suppl_tabs$cohorts) %>% 
  set_widths(c(2.7, rep(2.3, 6), 2.3, 2)) %>% 
  footnote(1, 8:9, 
           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

flextable::flextable(suppl_tabs$sub_clinic) %>% 
  merge_v(1) %>% 
  set_widths(c(2.2, 2.7, rep(2.8, 4), 2.3, 2)) %>% 
  footnote(1, 7:8, 
           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

```{r tab-fgf-genetics, tab.cap = "Frequency of somatic mutations and copy number variants of FGF-, FGFR-, and FGFBP-coding genes in the investigated cohorts. Statistical significance for differences between the cohorts was assessed by chi-squared test with Cramer's V effect size statistic. P values were corrected for multiple testing with the false discovery rate method. Frequency of alterations found in at least 1% of specimens in the BCAN, GENIE BLCA, IMvigor, MSK IMPACT, and TCGA BLCA are presented."}

flextable::flextable(suppl_tabs$genetic_alterations) %>% 
  merge_v(1) %>% 
  merge_v(2) %>% 
  merge_v(3) %>% 
  italic(j = 3) %>% 
  set_widths(c(2.5, 3, 2.2, 3, rep(2.5, 3))) %>% 
  theme_vanilla

@@

```{r tab-sub-genetics, tab.cap = "Frequency of somatic mutations and copy number variants of FGFR3 and FGF3/4/19 genes in the luminal papillary (LumP), luminal unstable (LumU), stoma-rich, and basal/squamous like (Ba/Sq) consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts. The frequencies are presented as percentages of the consensus class. Statistical significance of differences between the consensus classes was determined by weighted permutation test with Cramer's V effect size statistic. P values were corrected for multiple testing with the false discovery rate method."}

flextable::flextable(suppl_tabs$sub_genetics) %>% 
  merge_v(1:2, combine = TRUE) %>% 
  italic(j = 2) %>% 
  set_widths(c(2.3, 2.4, 2.3, rep(2.4, 3), rep(2.4, 2))) %>% 
  theme_vanilla

@@

```{r tab-expression, tab.cap = "Expression of FGFR-, FGF-, and FGFBP-coding genes at mRNA level was quantified in the TCGA BLCA, IMvigor, and BCAN cohort by RNA sequencing. The mRNA amounts were provided as log2-transformed gene counts (TCGA BLCA and IMvigor) or Z-scores of log2 gene counts (BCAN). Protein expression of the genes of interest was measured in the Human Protein Atlas (HPA) by immunohistochemistry (IHC). The protein amounts are metered by the IHC score defined as a product of staining intensity and quantity of positive cells. Median mRNA and protein amounts with interquartile ranges, ranges, and numbers of cancer samples are listed."}

flextable::flextable(suppl_tabs$expression) %>% 
  italic(i = 3:47, j = 1) %>% 
  set_widths(c(2.5, rep(3.5, 5))) %>% 
  theme_vanilla

@@

```{r tab-sub-dge, tab.cap = "Expression of FGF-, FGFR-, and FGFBP-coding genes in the consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts. Statistical significance of differences between the molecular classes was determined by one-way ANOVA with eta-square effect size statistic. P values were corrected for multiple testing with the false discovery rate method. log2-transformed expression levels are presented as medians with interquartile ranges and ranges for the top genes whose expression differentiated between the molecular consensus classes (top 15 largest eta-square values shared by all cohorts)."}

flextable::flextable(suppl_tabs$sub_dge) %>% 
  merge_v(1) %>% 
  italic(j = 2, italic = suppl_tabs$sub_dge$Variable != 'Samples, N') %>% 
  set_widths(c(1.6, 2.1, rep(3.2, 4), rep(2.3, 2))) %>% 
  theme_vanilla

@@

```{r tab-sub-sig, tab.cap = "Single sample gene set enrichment analysis (ssGSEA) scores of FGFR-related gene signatures in the consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts. Statistical significance of differences between the molecular classes was determined by one-way ANOVA with eta-square effect size statistic. P values were corrected for multiple testing with the false discovery rate method. ssGSEA scores are presented as medians with interquartile ranges and ranges for the top signatures whose scores differentiated between the molecular consensus classes (top 15 largest eta-square values shared by all cohorts)."}

flextable::flextable(suppl_tabs$sub_sig) %>% 
  merge_v(1) %>% 
  set_widths(c(1.6, 3.2, rep(2.9, 4), rep(2.3, 2))) %>% 
  theme_vanilla

@@

```{r tab-shap, tab.cap = "Importance of explanatory variables for the Elastic Net and Random Forest models of MIBC consensus molecular classes estimated by the SHAP algorithm. The models employed the FGFR-, FGF-, and FGFBP-coding genes as explanatory factors. Contribution of the explanatory factors to predictions of consensus molecular classes was assessed by mean absolute values of SHAP metrics (Shapley additive explanations). Top five most influential genes for the predictions of luminal papillary (LumP), luminal unstable (LumU), stroma-rich, and basal/squamous-like (Ba/Sq) classes by the Elastic Net and Random Forest algorithms are listed."}
